\cursedforest\ - A Random Forest Implementation for ``Big'' and ``Wide'' Data \\ Supplementary Information

### 2017-04-25 doc/publications/plos16/nm.tex
changes to nm.tex
1) million not Million
2) removed the spacing symbold from  \cursedforest\  -- the definition ends in an \xspace which will space it properly
3) fixed double quotes so LaTeX  will draw left and right quotation marks
4) make n and p italic
5) we have inconsistent capitalization on Bone Mineral Density, Bone mineral density, bone mineral density
I think we get rid of all the capitals
6) datasets or data sets ?
7)  We have a number of instances where we start a sentence with a reference. This is OK in an author-year style but with a numeric style I have had editors complain abou

Analyses are performed on ever larger datasets with unprecedentedly detailed information per data point. Specifically in genomics, cohort sizes are in the thousands, each comprising of Millions of variants that make up a person's genetic differences. Traditional approaches are unable to cater for this task and even standard implementations using novel compute paradigms, such as Hadoop/Spark, are limited by the finite resource of computer memory on such data. Specifically, Spark ML's supervised machine learning methods suffer from "the curse of dimensionality". That is, they scale well with samples (n) but not with features (p, variants).

We introduce \cursedforest, a tailored implementation of random forests, designed to handle data with extremely large number of variables per sample. We showcase \cursedforest\ to accurately predict ethnicity from whole genome sequencing profiles of 2,500 individuals and 80 million genomic variants each (1000 Genomes Project) in 30 minutes. \cursedforest\ was the only implementation scaling to the full dataset, outperforming other methods on smaller subsets by 80\% or more. We also demonstrate a GWAS-style analysis using  \cursedforest's feature selection approach on an imputed SNP array dataset (2000 x 7 Million) on Bone mineral density (BMD), successfully predicting known BMD-genes as well as identifying novel candidates.

\cursedforest is included in our earlier variant interpretation package, VariantSpark, allowing it to perform near real-time classification of population-scale patient cohorts in "patients like mine" scenarios, as well as performing GWAS analysis on large unfiltered whole genome sequencing cohorts.

@@

Here we introduce \cursedforest, a tailored Hadoop/Spark-based implementation of random forests specifically designed to cater
for ``big'' (many samples) and ``wide'' (many features) datasets. 
\cursedforest extends our previously developed variant interpretation framework, VariantSpark~\cite{OBrien2015}, to now offer
supervised as well as unsupervised ML algorithms in the Spark framework.
In our implementation a Spark application runs on a ``driver'' node and
distributes tasks to many ``worker'' nodes, or ``executors''. 
By also utilising Spark to read in and manipulate the standard genomic variant format (VCF)
directly, \cursedforest outperforms existing tools even on small datasets where multithreading generally performs
well. Harnessing the virtually unlimited capability to parallelise tasks, \cursedforest can hence explore the solution 
space faster by building a larger number of diverse models to generate a consensus from.

@@

The benefits of random forest models are:
\begin{itemize}
\item they give an out-of-bag (OOB) estimate of model accuracy, by using the samples not selected for the bootstrapped sample
  as an independent test set for each tree;
\item they provide a measure of variable importance, for example by summing the number of times a variable is used to
  split a node, over all nodes and all trees. More complex measures of variable importance are also used;
\item they inherit some of the qualities of tree models (but not, for example, the handling of missing values via surrogate
  variables. This would be difficult as the trees are split on random subsets of variables).
\end{itemize}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% This section talks about the datasets
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Datasets}
\subsubsection{1000 Genomes Project}
We obtained the 1000 Genomes Project data as VCF files from their FTP site.  Each VCF file contains the variants for
every individual across one chromosome.  We use the phase 3 dataset which contains 2,504 individuals with approximately
81 million variant positions (i.e. a matrix of $2,504 \times 81$ million).  For compatibility with the majority of the
algorithms we demonstrate, we convert the original variant representations to double representations.  For example, we
store the absence of a variant (0\textbar0) as 0, a heterozygous
variant (1\textbar0 or 0\textbar1) as 1, and a
homozygous variant (1\textbar1) as 2.  However, this is not the only possible encoding \cite{Goldstein.et.al.2011}. Note that from
the encodings listed in Table~\ref{table:encodings}, the additive encoding will present the most difficulty for a random
forest model.

@@

The model parameters are $w_i = 1/\sqrt{2^{i-1}}$ for $i =1,\ldots 5$ and we set 
\begin{equation}
\label{equation:synthetic.data}
z =  \sum_{i=1}^{i=5} {w_i x_i}.
\end{equation}
We let $\sigma_\epsilon^2 = Var(z)(1-\theta)/\theta $  where $\theta$ is a parameter controlling the fraction of variance explained by the informative variables and in our study we chose $\theta = 0.125$ . Then 
$y =   z + \epsilon$ where $\epsilon \sim  N(0, \sigma_\epsilon^2).$
The dichotomous response is generated by thresholding $y$ at the $0.5$ quantile.

$$
\ddot{y} =    \left\{ \begin{array}{rcl}
			  0 & \mbox{for} & y \leq Q2(y) \\
			  1 & \mbox{for} & y > Q2(y)
		 \end{array}\right.
$$

\subsection{Parameter settings}
We consider the parameter settings for the random forest algorithm. We use the R notation from the \randomforest
package \cite{Liaw.and.Weiner.2002} which incorporates the original Fortran code by Brieman and Cutler.  We incorporate
the advice of \cite{Liaw.and.Weiner.2002}, which we have found mirrors our own experience:
\begin{itemize}
\item \ntree{}  -- the number of trees.  The number of trees necessary for good performance grows with the number of
  predictors.  \cite{Liaw.and.Weiner.2002} suggest that a high \ntree is necessary to get stable estimates of variable
  importance and proximity; however, even though the variable importance measures may vary from run to run, we note that
  it is possible for a random forest model to have a poorer fit and still have an accurate ranking of variable
  importance;
\item \mtry{}  -- the number of variables considered at each split (if \mtry=$p$, we have a boosted decision
  tree model).  If one has a very large number of variables but expects only very few to be ``important'', using larger \mtry may give
  better performance;
\item the size and complexity of the individual trees is controlled in \randomforest by setting \texttt{nodesize}, the
  minimum size of terminal nodes. It is controlled in Spark ML by setting \texttt{maxDepth}, the maximum depth of each
  tree in the forest.\footnote{The Spark ML documentation \cite{Spark.2016} sets \texttt{maxDepth} to 4 in their
    classification example. This may give an estimate with a higher bias and, as such, may be a poor choice. See the
    discussion in \cite{Dietterich.2002} where it is shown that bagging may reduce bias as well as variance but is more
    complicated than first thought. They show that bagging may reduce bias as well as variance but give a    
    better result for low bias learners. \cite{Dietterich.2002} notes that if the bootstrap replicate approximation were
    correct (i.e.~if the bootstrap sample came from an identical distribution to the data), then bagging would reduce
    variance without changing bias.}
\end{itemize}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Section 2: Feature Selection on real data 
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\subsection{GWAS-style analysis using \cursedforest\ identifies novel variants associated with Bone Mineral Density}
\subsection{GWAS-style analyses using \cursedforest\ faithfully identify known disease-accociated genes}
In this section we use \cursedforest\ to perform feature selection on over 7.2 million genomic variants and identify the
genomic locations associated with Bone Mineral Density (BMD)~\cite{Duncan.2011}. 
%We build a model on the BMD dataset as described in Methods using a binary high or low BMD value as the label.  
We rank variants by their Gini impurity, with the most important one ranked at position 1, and return the top 100,000 variants.

To test whether the associated variants have biological meaning, we investigate if variants in or near
BMD-associated genes are enriched toward the top of our ranked list. To form the list of "known genes", we
combine the 26 known and 6 novel BMD genes listed by Duncan {\it et al.}~\cite{Duncan.2011} with the genes identified by
Estrada {\it et al.}~\cite{Estrada2012} and Lui {\it et al.}~\cite{Liu2008}, resulting in 102 genes.  All genes are covered within the
top 40,000 variants and a gene name is assigned to the variant if it falls within 100 kb of this gene.

@@

\subsection{\cursedforest\ amplifies signal over traditional methods to identify novel associations}

Figure 1b shows a traditional GWAS analysis using logistic regression~\cite{Duncan.2011} where the resulting $p$-values do not clearly distinguish these known BMD genes from the noise. 
In contrast, \cursedforest\ seems to amplify the signal to identify associated features (Figure 1a). 
As such, this approach allows to employ a very stringent cut-off for identifying strongly associated feature. Here we chose a 99.98 False Discovery Rate (FDR) from 60 random experiments (see Methods).

@@

Above this FDR cut-off we identify four novel locations as annotated in Figure 1a.
%~\ref{figure:rbo-prod.png}. 
Most notably a lincRNA (LINC01006) on chromosome 7, which has been associated with serum creatinine levels in two independent GWAS studies~\cite{Pattaro2010,Chambers2010}. 
Subjects with low serum creatinine levels were at a higher risk for low BMD~\cite{Huh2015} and in a rat study high creatine monohydrate diet resulted in greater lumbar BMD~\cite{Antolic2007}, 
with creatinine being the breakdown product of creatine. The variants overlapping the 5' end of the lincRNA are the highest ranked features in this region and all LD SNPs also show high importance (see Fig 3d). Also amonst the top 30 features is PRODH (proline dehydrogenase) located on 22q11, a loci observed by Stagi {\it
  et al.}~\cite{Stagi.2010} to significantly reduce bone mass in individuals with a micro-deletion in this region. 
PRODH catalyses the first step in proline degradation, and as proposed by Phang {\it et al.}~\cite{Phang.2008}, proline is made available by collagen degradation. 
Again, all LD SNPs also show high importance (see Figure 3c). %~\ref{zoomplot}). 
%Ranked at position 5 (rs6961378) and 18 (rs146034056) are located on chromosome 7 overlapping the lincRNA (LINC01006), which as been linked to vertebrate development~\cite{Woolfe2005}.
In contrast rs17195090 and rs112558362 are single feature loci (Supplementary Figure 4) hence likely due to erroneous imputation or genotyping.

@@

%\cursedforest\ can successfully perform GWAS-type analyses replicating previous findings and uncovering novel associations due to random forests also modelling interacting features.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Section 3: Classification selection
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{\cursedforest outperforms existing methods for classification problems}
In this section we demonstrate \cursedforest's classification accuracy on biological data and compare its performance against other published methods. 
We do this by predicting the ethnicity of the individuals in the 1000 Genomes Project from their genomic profiles and evaluate against the annotated label. 
We record the prediction error (OOB) as well as the runtime for each method. 
%@Aidan is 12 correct?
We limit the number of executors to 12 that are available to \cursedforest\ to keep the comparison fair for all methods.

As shown in Fig.~\ref{figure:1000genomes}, only \cursedforest\ is able to scale to the full phase3\_chr1-22 dataset (2,504 samples x 19,328,051 variants), with
the standard random forest implementation in R running out of resources on the relatively low-dimensional
phase1\_chr22 dataset (1,092 x 490,036). 
Ranger's c++ implementation and the R front-end version of it are able to process the second largest dataset (phase3\_chr1-3 with 2,504 x 19,328,051), however, both are unable to process larger datasets successfully. 
While the c++ implementation runs out of memory the R front-end, which also covers the loading and pre-processing, needs an unreasonable compute time for this task (>20h), see Supplementary Table 1.
The standard implementation of random forest in SparkML already runs out of memory after processing the third largest dataset, phase3\_chr1 (2,504 x 6,450,364). 
This demonstrates that Spark ML, although designed to handle many samples, is unable to efficiently cope with large numbers of features.

All tested methods have a worse runtime than \cursedforest\ for the datasets they processed successfully, except for the smallest dataset where the overhead of Spark's distributed platform made ranger's R implementation faster with 1 min 10 seconds compared to 1 min 50 seconds. For example, on phase3\_chr1-3 \cursedforest\ finishes in 1 hour, with ranger(R) requiring 4 hours and SparkML already needing 8 hours on a smaller dataset. 
Extrapolating from this, all comparable implementations hence would require more than 24 hours to finish, assuming the use of larger memory nodes.

As our implementation can efficiently utilise large numbers of commodity nodes, the performance of \cursedforest\ can be substantially improved further. 
%@Aidan is 130 correct?
Table~\ref{cursedforesttable} shows the runtime when utilising the full Spark cluster (130 worker nodes).

Building a model with \cursedforest\ on phase3\_chr1, which previously took 20 minutes, can be reduced to 4 minutes. 
Subsequently, phase3\_chr1-3 takes about 11 minutes, and building a
model on the entire genome is 37 minutes.

Not only does the job complete successfully using \cursedforest\, but building a random
forest model on all 22 autosomes (phase3\_chr1-22)  leads to a more accurate result. 
The error can be reduced from OOB=0.04 when using only data from Chr1, which represents processing power limit for most tools, down to OOB=0.01 when using the full dataset.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Section 1:  Feature Selection and classification on synthetic data
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{\cursedforest can evaluate variable importance and perform classification for millions of features}
%\label{section:synthetic}
In this section we test the limits of \cursedforest\'s association testing and classify ability on high-dimensional data. 
%In this section we demonstrate the two features of \cursedforest, namely: variable selection analysis to select the most
%associated set of features amongst millions of dimensions; and its ability to accurately classify the labels of
%high-dimensional samples.  
We therefore generate a synthetic dataset with 2.5 million variables ($p=2,500,000$), of which 5 
are related to the response variable, and 5000 samples ($n=5000$), as discussed in the methods section.

@@

As shown in Fig~\ref{figure:out-of-bag-prediction-error-prod.png} the default value for \mtry (number of variables
selected at each node) does not result in a good classification performance for this large feature dataset. Note that
the plot shows the proportion, $n/p,$ for \mtry\ with the default being $\sqrt{p} \approx 1581$ ($n/p=0.0006$).  The
OOB for this value of \mtry does not drop below 0.5, even when the number of trees are increased (\ntree).

Increasing \mtry\ in combination with \ntree\ yields the best performance with the OOB error essentially constant around 0.4
across a large range of of \mtry and \ntree values.  This is in contrast to the feature-selection performance, where the
RBO measure heavily depends on \ntree\ and gives better results with lower values of \mtry.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Section 4: Sensitivity to different parameters and scalability
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Scalability analysis}
In this section we explore the performance of \cursedforest\ in more detail by testing its ability to scale to different sizes of data
and computational resources.

@@

\begin{figure}[tbhp]
  \begin{adjustwidth}{-1.00in}{0in}
    \caption{\textbf{Scalability of the wide random forest on the synthetic dataset of 2.5M features and 5K samples.}}
    \label{figure:synthetictiming}
    \begin{subfigure}[b]{0.45\linewidth}
      \centering
      \includegraphics[totalheight=6cm]{./figs/mtry_cpu.png} 
      \caption{Time in seconds to build 100 trees for different \mtry\ fractions. } 
      \label{figure:synthetictiming.a} 
      \vspace{4ex}
    \end{subfigure} 
    \hfill
    \begin{subfigure}[b]{0.45\linewidth}
      \centering
      \includegraphics[totalheight=6cm]{./figs/cpu_mtry_trees_per_hour.png}
      \caption{Number of trees build per hour when using a growing number of CPU cores.}
      \label{figure:synthetictiming.b}
      \vspace{4ex}
    \end{subfigure} 
  \end{adjustwidth}
\end{figure}

@@

% no data in this table 
% \begin{table}[!ht]
% \caption{
% {\bf Sensitivity analysis on parameter choice.}}
% \begin{tabular}{|l|l|l|l|l|l|l|l|}
% \hline
% %\multicolumn{4}{|l|}{\bf Heading1} & \multicolumn{4}{|l|}{\bf Heading2}\\ \hline
% \bf{ \mtry\ }  & \bf{Mtree} & \bf{depth} & \bf{Accuracy} & \bf{Runtime} & \bf{Memory} \\
% \hline
% &&&&&\\ \hline
% \end{tabular}
% \begin{flushleft} 
%   Table notes here.
% \end{flushleft}
% \label{table2}
% \end{table}

@@

To showcase the breadth of application scope, we also demonstrate that \cursedforest's implementation aids in classification problems such as predicting the ethnicity of the 1000 genome project. 
By comparing \cursedforest\ to other implementations (including those optimised for large datasets) we have demonstrated that runtime can be improved by 80\% allowing analysis in timeframes relevant for point-of-care diagnostics such as for "Patients like mine" scenarios thereby enabling evidence based self learning health care.

In summary, \cursedforest\  offers the broadly needed capability of performing sophisticated machine learning tasks on big, complex life science data. Specifically, the here demonstrated ability to uncover novel biological information from previously analysed GWAS studies showcases the enormous potential of re-analysing the GWAS catalogue given this unprecedented computational capacity to perform joint-loci association testing and identify the much-anticipated cumulative multi-gene disease contribution.

>>>>

Analyses are being performed on ever larger datasets with unprecedentedly detailed information per data point. Specifically in
genomics, cohort sizes are in the thousands, each comprising of millions of variants that make up a person's genetic
differences. Traditional approaches are unable to cater for this task and even standard implementations using novel
compute paradigms, such as Hadoop/Spark, are limited by the finite resource of computer memory on such
data. Specifically, Spark ML's supervised machine learning methods suffer from ``the curse of dimensionality''. That is,
they scale well with samples ($n$) but not with features ($p$, variants).

We introduce \cursedforest, a tailored implementation of random forests, designed to handle data with extremely large
number of variables per sample. We showcase \cursedforest to accurately predict ethnicity from whole genome sequencing
profiles of 2,500 individuals and 80 million genomic variants each (1000 Genomes Project) in 30 minutes. \cursedforest
was the only implementation scaling to the full dataset, outperforming other methods on smaller subsets by 80\% or
more. We also demonstrate a GWAS-style analysis using \cursedforest's feature selection approach on an imputed SNP array
dataset (2000 x 7 Million) on Bone mineral density (BMD), successfully predicting known BMD-genes as well as identifying
novel candidates.

\cursedforest is included in our earlier variant interpretation package, VariantSpark, allowing it to perform near
real-time classification of population-scale patient cohorts in ``patients like mine'' scenarios, as well as performing
GWAS analysis on large unfiltered whole genome sequencing cohorts.

@@

Here we introduce \cursedforest, a tailored Hadoop/Spark-based implementation of random forests specifically designed to cater
for ``big'' (many samples) and ``wide'' (many features) datasets. 
\cursedforest extends our previously developed variant interpretation framework, VariantSpark~\cite{OBrien2015}, to now offer
supervised as well as unsupervised ML algorithms in the Spark framework.
In our implementation a Spark application runs on a ``driver'' node and
distributes tasks to many ``worker'' nodes, or ``executors''. 
By also utilising Spark to read in and manipulate the standard genomic variant format (VCF)
directly, \cursedforest outperforms existing tools even on small datasets where multithreading generally performs
well. Harnessing the virtually unlimited capability to parallelise tasks, \cursedforest can hence explore the solution 
space faster by building a larger number of diverse models to generate a consensus classifier.

@@

The benefits of random forest models are:
\begin{itemize}
\item they give an out-of-bag (OOB) estimate of model accuracy, by using the samples not selected for the bootstrapped sample
  as an independent test set for each tree;
\item they provide a measure of variable importance, for example by summing the number of times a variable is used to
  split a node, over all nodes and all trees. Other, more complex, measures of variable importance are also used;
\item they inherit some of the qualities of tree models (but not, for example, the handling of missing values via surrogate
  variables. This would be difficult as the trees are split on random subsets of variables).
\end{itemize}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% This section talks about the datasets
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Datasets}
\subsubsection{1000 Genomes Project}
We obtained the 1000 Genomes Project data as VCF files from their FTP site.  Each VCF file contains the variants for
every individual across one chromosome.  We use the phase 3 dataset which contains 2,504 individuals with approximately
81 million variant positions (i.e. a matrix of $2,504 \times 81$ million).  For compatibility with the majority of the
algorithms we demonstrate, we convert the original variant representations to double representations.  For example, we
store the absence of a variant (0\textbar0) as 0, a heterozygous
variant (1\textbar0 or 0\textbar1) as 1, and a
homozygous variant (1\textbar1) as 2, so we have an ``additive'' model.  However, this is not the only possible encoding \cite{Goldstein.et.al.2011}. Note that from
the encodings listed in Table~\ref{table:encodings}, the additive encoding will present the most difficulty for a random
forest model.

@@

The model parameters are $w_i = 1/\sqrt{2^{i-1}}$ for $i =1,\ldots 5$ and we set 
\begin{equation}
\label{equation:synthetic.data}
z =  \sum_{i=1}^{i=5} {w_i x_i}.
\end{equation}
We let $\sigma_\epsilon^2 = Var(z)(1-\theta)/\theta $  where $\theta$ is a parameter controlling the fraction of variance explained by the informative variables and in our study we chose $\theta = 0.125$ . Then 
$y =   z + \epsilon$ where $\epsilon \sim  N(0, \sigma_\epsilon^2).$
The dichotomous response is generated by thresholding $y$ at the $0.5$ quantile,

$$
\ddot{y} =    \left\{ \begin{array}{rcl}
			  0 & \mbox{for} & y \leq Q2(y) \\
			  1 & \mbox{for} & y > Q2(y).
		 \end{array}\right 
$$

\subsection{Parameter settings}
We consider the parameter settings for the random forest algorithm. We use the R notation from the \randomforest
package \cite{Liaw.and.Weiner.2002} which incorporates the original Fortran code by Brieman and Cutler.  We incorporate
the advice of \cite{Liaw.and.Weiner.2002}, which we have found mirrors our own experience:
\begin{itemize}
\item \ntree{}  -- the number of trees.  The number of trees necessary for good performance grows with the number of
  predictors.  \cite{Liaw.and.Weiner.2002} suggest that a high \ntree is necessary to get stable estimates of variable
  importance and proximity; however, even though the variable importance measures may vary from run to run, we note that
  it is possible for a random forest model to have a poorer fit and still have an accurate ranking of variable
  importance;
\item \mtry{}  -- the number of variables considered at each split (if \mtry=$p$, we have a boosted decision
  tree model).  If one has a very large number of variables but expects only very few to be ``important'', using larger \mtry may give
  better performance;
\item the size and complexity of the individual trees is controlled in \randomforest by setting \texttt{nodesize}, the
  minimum size of terminal nodes. It is controlled in Spark ML by setting \texttt{maxDepth}, the maximum depth of each
  tree in the forest.\footnote{The Spark ML documentation \cite{Spark.2016} sets \texttt{maxDepth} to 4 in their
    classification example. This may give an estimate with a higher bias and, as such, may be a poor choice. See the
    discussion in \cite{Dietterich.2002} where it is shown that the action of a RF is complicated than first
    thought. They show that bagging may reduce bias as well as variance but give a better result for low bias
    learners. \cite{Dietterich.2002} notes that if the bootstrap replicate approximation were correct (i.e.~if the
    bootstrap sample came from an identical distribution to the data), then bagging would reduce variance without
    changing bias.}
\end{itemize}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Section 2: Feature Selection on real data 
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\subsection{GWAS-style analysis using \cursedforest identifies novel variants associated with Bone Mineral Density}
\subsection{GWAS-style analyses using \cursedforest faithfully identify known disease-accociated genes}
In this section we use \cursedforest to perform feature selection on over 7.2 million genomic variants and identify the
genomic locations associated with Bone Mineral Density (BMD)~\cite{Duncan.2011}. 
%We build a model on the BMD dataset as described in Methods using a binary high or low BMD value as the label.  
We rank variants by their Gini impurity, with the most important one ranked at position 1, and return the top 100,000 variants.

To test whether the associated variants have biological meaning, we investigate if variants in or near
BMD-associated genes are enriched toward the top of our ranked list. To form the list of ``known genes'', we
combine the 26 known and 6 novel BMD genes listed by Duncan {\it et al.}~\cite{Duncan.2011} with the genes identified by
Estrada {\it et al.}~\cite{Estrada2012} and Lui {\it et al.}~\cite{Liu2008}, resulting in 102 genes.  All genes are covered within the
top 40,000 variants and a gene name is assigned to the variant if it falls within 100 kb of this gene.

@@

\subsection{\cursedforest amplifies signal over traditional methods to identify novel associations}

Figure 1b shows a traditional GWAS analysis using logistic regression~\cite{Duncan.2011} where the resulting $p$-values do not clearly distinguish these known BMD genes from the noise. 
In contrast, \cursedforest seems to amplify the signal to identify associated features (Figure 1a). 
As such, this approach allows to employ a very stringent cut-off for identifying strongly associated feature. Here we chose a 99.98 False Discovery Rate (FDR) from 60 random experiments (see Methods).

@@

Above this FDR cut-off we identify four novel locations as annotated in Figure 1a.
%~\ref{figure:rbo-prod.png}. 
Most notably a lincRNA (LINC01006) on chromosome 7, which has been associated with serum creatinine levels in two independent GWAS studies~\cite{Pattaro2010,Chambers2010}. 
Subjects with low serum creatinine levels were at a higher risk for low BMD~\cite{Huh2015} and in a rat study high creatine monohydrate diet resulted in greater lumbar BMD~\cite{Antolic2007}, 
with creatinine being the breakdown product of creatine. The variants overlapping the 5' end of the lincRNA are the highest ranked features in this region and all LD SNPs also show high importance (see Fig 3d). Also amongst the top 30 features is PRODH (proline dehydrogenase) located on 22q11, a loci observed by Stagi {\it
  et al.}~\cite{Stagi.2010} to significantly reduce bone mass in individuals with a micro-deletion in this region. 
PRODH catalyses the first step in proline degradation, and as proposed by Phang {\it et al.}~\cite{Phang.2008}, proline is made available by collagen degradation. 
Again, all LD SNPs also show high importance (see Figure 3c). %~\ref{zoomplot}). 
%Ranked at position 5 (rs6961378) and 18 (rs146034056) are located on chromosome 7 overlapping the lincRNA (LINC01006), which as been linked to vertebrate development~\cite{Woolfe2005}.
In contrast rs17195090 and rs112558362 are single feature loci (Supplementary Figure 4) hence likely due to erroneous imputation or genotyping.

@@

%\cursedforest can successfully perform GWAS-type analyses replicating previous findings and uncovering novel associations due to random forests also modelling interacting features.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Section 3: Classification selection
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{\cursedforest outperforms existing methods for classification problems}
In this section we demonstrate \cursedforest's classification accuracy on biological data and compare its performance against other published methods. 
We do this by predicting the ethnicity of the individuals in the 1000 Genomes Project from their genomic profiles and evaluate against the annotated label. 
We record the prediction error (OOB) as well as the runtime for each method. 
%@Aidan is 12 correct?
We limit the number of executors to 12 that are available to \cursedforest to keep the comparison fair for all methods.

As shown in Fig.~\ref{figure:1000genomes}, only \cursedforest is able to scale to the full phase3\_chr1-22 dataset (2,504 samples x 19,328,051 variants), with
the standard random forest implementation in R running out of resources on the relatively low-dimensional
phase1\_chr22 dataset (1,092 x 490,036). 
Ranger's c++ implementation and the R front-end version of it are able to process the second largest dataset (phase3\_chr1-3 with 2,504 x 19,328,051), however, both are unable to process larger datasets successfully. 
While the c++ implementation runs out of memory the R front-end, which also covers the loading and pre-processing, needs an unreasonable compute time for this task (>20h), see Supplementary Table 1.
The standard implementation of random forest in SparkML already runs out of memory after processing the third largest dataset, phase3\_chr1 (2,504 x 6,450,364). 
This demonstrates that Spark ML, although designed to handle many samples, is unable to efficiently cope with large numbers of features.

All tested methods have a worse runtime than \cursedforest for the datasets they processed successfully, except for the smallest dataset where the overhead of Spark's distributed platform made ranger's R implementation faster with 1 min 10 seconds compared to 1 min 50 seconds. For example, on phase3\_chr1-3 \cursedforest finishes in 1 hour, with ranger(R) requiring 4 hours and SparkML already needing 8 hours on a smaller dataset. 
Extrapolating from this, all comparable implementations hence would require more than 24 hours to finish, assuming the use of larger memory nodes.

As our implementation can efficiently utilise large numbers of commodity nodes, the performance of \cursedforest can be substantially improved further. 
%@Aidan is 130 correct?
Table~\ref{cursedforesttable} shows the runtime when utilising the full Spark cluster (130 worker nodes).

Building a model with \cursedforest on phase3\_chr1, which previously took 20 minutes, can be reduced to 4 minutes. 
Subsequently, phase3\_chr1-3 takes about 11 minutes, and building a
model on the entire genome is 37 minutes.

Not only does the job complete successfully using \cursedforest, but building a random
forest model on all 22 autosomes (phase3\_chr1-22)  leads to a more accurate result. 
The error can be reduced from OOB=0.04 when using only data from Chr1, which represents processing power limit for most tools, down to OOB=0.01 when using the full dataset.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Section 1:  Feature Selection and classification on synthetic data
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{\cursedforest can evaluate variable importance and perform classification for millions of features}
%\label{section:synthetic}
In this section we test the limits of \cursedforest's association testing and classify ability on high-dimensional data. 
%In this section we demonstrate the two features of \cursedforest, namely: variable selection analysis to select the most
%associated set of features amongst millions of dimensions; and its ability to accurately classify the labels of
%high-dimensional samples.  
We therefore generate a synthetic dataset with 2.5 million variables ($p=2,500,000$), of which 5 
are related to the response variable, and 5000 samples ($n=5000$), as discussed in the methods section.

@@

As shown in Fig~\ref{figure:out-of-bag-prediction-error-prod.png} the default value for \mtry (number of variables
selected at each node) does not result in a good classification performance for this large feature dataset. Note that
the plot shows the proportion, $n/p,$ for \mtry with the default being $\sqrt{p} \approx 1581$ ($n/p=0.0006$).  The
OOB for this value of \mtry does not drop below 0.5, even when the number of trees are increased (\ntree).

Increasing \mtry in combination with \ntree yields the best performance with the OOB error essentially constant around 0.4
across a large range of of \mtry and \ntree values.  This is in contrast to the feature-selection performance, where the
RBO measure heavily depends on \ntree and gives better results with lower values of \mtry.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Section 4: Sensitivity to different parameters and scalability
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Scalability analysis}
In this section we explore the performance of \cursedforest in more detail by testing its ability to scale to different sizes of data
and computational resources.

@@

\begin{figure}[tbhp]
  \begin{adjustwidth}{-1.00in}{0in}
    \caption{\textbf{Scalability of the wide random forest on the synthetic dataset of 2.5M features and 5K samples.}}
    \label{figure:synthetictiming}
    \begin{subfigure}[b]{0.45\linewidth}
      \centering
      \includegraphics[totalheight=6cm]{./figs/mtry_cpu.png} 
      \caption{Time in seconds to build 100 trees for different \mtry fractions. } 
      \label{figure:synthetictiming.a} 
      \vspace{4ex}
    \end{subfigure} 
    \hfill
    \begin{subfigure}[b]{0.45\linewidth}
      \centering
      \includegraphics[totalheight=6cm]{./figs/cpu_mtry_trees_per_hour.png}
      \caption{Number of trees build per hour when using a growing number of CPU cores.}
      \label{figure:synthetictiming.b}
      \vspace{4ex}
    \end{subfigure} 
  \end{adjustwidth}
\end{figure}

@@

% no data in this table 
% \begin{table}[!ht]
% \caption{
% {\bf Sensitivity analysis on parameter choice.}}
% \begin{tabular}{|l|l|l|l|l|l|l|l|}
% \hline
% %\multicolumn{4}{|l|}{\bf Heading1} & \multicolumn{4}{|l|}{\bf Heading2}\\ \hline
% \bf{ \mtry }  & \bf{Mtree} & \bf{depth} & \bf{Accuracy} & \bf{Runtime} & \bf{Memory} \\
% \hline
% &&&&&\\ \hline
% \end{tabular}
% \begin{flushleft} 
%   Table notes here.
% \end{flushleft}
% \label{table2}
% \end{table}

@@

To showcase the breadth of application scope, we also demonstrate that \cursedforest's implementation aids in classification problems such as predicting the ethnicity of the 1000 genome project. 
By comparing \cursedforest to other implementations (including those optimised for large datasets) we have demonstrated that runtime can be improved by 80\% allowing analysis in timeframes relevant for point-of-care diagnostics such as for "Patients like mine" scenarios thereby enabling evidence based self learning health care.

In summary, \cursedforest  offers the broadly needed capability of performing sophisticated machine learning tasks on big, complex life science data. Specifically, the here demonstrated ability to uncover novel biological information from previously analysed GWAS studies showcases the enormous potential of re-analysing the GWAS catalogue given this unprecedented computational capacity to perform joint-loci association testing and identify the much-anticipated cumulative multi-gene disease contribution.
